\documentclass[12pt,platino]{article}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}

\title{Assignment: Data Infrastructure at Meta \\ CSE 817: Data Engineering}
\date{}

\begin{document}
	
	\maketitle
	
 
\textit{  Research and write a detailed note on the data infrastructure at Meta. You will use the link provided below as your primary resource and present your findings in a video presentation. Your submission should reflect your own understanding and analysis of the material.}
	
	\section*{Instructions}
	\begin{enumerate}
		\item \textbf{Research}: \\
		Visit the link: \href{https://engineering.fb.com/category/data-infrastructure/}{Meta Data Infrastructure} to explore Meta’s data infrastructure and understand the systems, technologies, and approaches used.
		
		\item \textbf{Written Note}: \\
		Write a comprehensive note summarizing the key aspects of Meta’s data infrastructure. Highlight specific tools, architectures, and strategies that Meta employs to manage, store, and analyze data at scale.
		
		\item \textbf{Video Presentation}: \\
		Create a 5–7 minute video summarizing your research findings. Your presentation should be clear, well-organized, and demonstrate a strong understanding of the topic.
		
		\item \textbf{Submission}: \\
		Share a URL to your video presentation. Ensure the link is accessible and that the video can be viewed by the instructor.
	\end{enumerate}
	
	\section*{Guidelines}
	\begin{itemize}
		\item The work must be your own. Do not rely on AI tools or external help to generate your content. Any form of plagiarism or dishonesty will result in penalties.
		\item Your note and video should reflect your original analysis and understanding of the material.
	\end{itemize}
	
 
\end{document}